Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2019-07-25 unimodular-covers.tex
wrong file added
went through Giulia's changes

\begin{abstract}
We show that the following classes of three-dimensional lattice polytopes have unimodular covers: the class of all parallelepipeds, the class of all centrally symmetric polytopes, and the class of all three-dimensional Cayley sums $\cayley(P,Q)$ where the normal fan of $Q$ refines that of $P$. This improves results of Beck et al.~(2018) and Haase et al.~(2008) where the last two classes were shown to be IDP.
\end{abstract}

@@

Oda (\cite{Oda1997}) posed several questions regarding smoothness and the IDP property for lattice polytopes.
%
\paco{I think this definition goes better here, and decided to "undeclare" it and the definition of weak Mink sum}
Following \cite{HaaseHof, Tsuchiya}, we say that a pair $(P, Q)$ of lattice polytopes has the integer decomposition property, or that \emph{the pair $(P,Q)$ is IDP}, if 
\begin{align*}
\label{eq:mixedIDP}
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\end{align*}
%
A lattice polytope $Q$ is called \emph{smooth} if it is simple and the primitive edge directions \giulia{changed this from "edge generators", which to me was unclear} at every vertex form a linear basis for the lattice; equivalently, if the projective toric variety defined by the normal fan of $Q$ is smooth. 
The following versions of Oda's questions are now considered conjectures~\cite{HNPS2008,mfo2007}, and they are open even in dimension three:
\begin{conjecture}
\label{conj:Oda}
\begin{enumerate}
\item 
\label{itm:smoothIDP}
(Related to problems 2 and 5 in \cite{Oda1997})
Every smooth lattice polytope is IDP.
\item 
\label{itm:mixedIDP}
(Related to problems 1, 3, 4, 6 in \cite{Oda1997}) Every pair $(P,Q)$ of lattice polytopes with $Q$ smooth and the normal fan of $Q$ refining that of $P$ is IDP.
%\begin{align}
%\label{eq:mixedIDP}
%(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
%\end{align}
\end{enumerate}
\end{conjecture}

When the normal fan of a $Q$ refines that of $P$, as in the second conjecture, we say that $P$ \emph{is a weak Minkowski summand of $Q$}, since this is equivalent to the existence of a polytope $P'$ such that $P+P' = k Q$ for some dilation constant $k>0$\paco{did not find this in the google books version of \cite{Grunbaum}; found it in Postnikov-Reiner-Williams, Tsuchiya, a survey by Rekha Thomas, and other places. Perhaps we can leave it wothout reference}. 
This property has the following algebraic implication for the projective toric variety $X_Q$: $P$ is a weak Minkowski summand of $Q$ if and only if the Cartier divisor on  $X_Q$ defined by $P$ is \emph{numerically effective}, or ``nef'' (see~\cite[Cor.~6.2.15, Prop.~6.3.12]{CLS}, but observe that there, what we call  ``weak Minkowski summand'' is called simply ``Minkowski summand''). 
%\paco{added this explanation}
%
%\begin{definition}
%Following \cite{HaaseHof, Tsuchiya} we say that the pair $(P, Q)$ has the integer decomposition property, or \emph{the pair $(P,Q)$ is IDP}, if 
%\begin{align*}
%(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
%\end{align*}
%\end{definition}
% 
%In this language,  \Cref{conj:Oda}\eqref{itm:mixedIDP} can be rephrased as: if $P$ is a weak Minkowski summand of a smooth polytope $Q$ then the pair $(P,Q)$ is IDP.

@@

The two-dimensional case of \Cref{conj:Oda}\eqref{itm:mixedIDP} is known to hold, with three different proofs by Fakhruddin~\cite{Fakhruddin}, Ogata~\cite{Ogata} and Haase et al.~\cite{HNPS2008}. This last one actually shows that smoothness of $Q$ is not needed. In dimension three, however, the conjecture fails without the smoothness assumption. Indeed, it fails if $P=Q$ is any non-unimodular \emph{empty tetrahedron}, by which we mean a lattice tetrahedron containing no lattice points other than its vertices (see the proof of \Cref{lemma:corner} for a classification of empty tetrahedra).
\paco{added def of empty tetrahedron}\giulia{I didn't like "details on them" and changed it to  "classification of empty tetrahedra". Is this ok?}

An alternative approach to \Cref{conj:Oda}\eqref{itm:mixedIDP} is via Cayley sums, which we discuss in  \Cref{sec:cayley}. 
Recall that the \emph{Cayley sum} of two lattice polytopes $Q_1,Q_2\subset \R^d$ is the lattice polytope
\[
\cayley(Q_1,Q_2) := \conv(Q_1\times\{0\} \cup Q_2 \times \{1\}) \subset \R^3.
\]
We normally require $\cayley(P,Q)$ to be full-dimensional (otherwise we can delete coordinates) but for this does not need $Q_1$ and $Q_2$ to be full-dimensional. It only requires the linear subspaces parallel to them to span $\R^d$.

@@

This has the following two consequences, also proved in \Cref{sec:cayley}.
A  \emph{prismatoid} is a polytope whose vertices all lie in two parallel facets. 
A polytope has width $1$ if its vertices lie in two \emph{consecutive} parallel lattice hyperplanes. Observe that this is the same as being ($SL(\Z,d)$-equivalent to) a Cayley sum.

@@

For each $i \in [4]$, let $T_i=\conv(q_i, p_j, p_k, p_l)$, with $\{i,j,k,l\}=[4]$; we call these $T_i$ the \emph{corner tetrahedra} of $C(T)$. Together with $T$ they triangulate of $C(T)$.
\end{definition}

The idea to have in mind is that of the regular tetrahedron inscribed in a cube, see \Cref{fig:circumscribed_parall}. 
Modulo an affine transformation this is exactly the situation of $T$ and $C(T)$.

\begin{figure}[htb]
\includegraphics[width=.7\textwidth]{circumscribed_parall}
\caption{In red we have the tetrahedron $T$, in black the circumscribed parallelepiped $C(T)$, and in blue the corner simplex $T_4$.}
\label{fig:circumscribed_parall}
\end{figure}
\giulia{I added a picture, now we can decide if we want it in the paper or not.}

@@

See \Cref{fig:corner}. Observe that $u+v=p_1+p_2=q_3+q_4$.
Now, this implies that the quadrilateral $\conv(p_1q_4p_2q_3)$ contains a fundamental domain for the lattice $\Z^2\times\{0\}$. Hence, its translate $\conv(q_2p_3q_1p_4)$ contains a fundamental domain for $\Z^2\times\{1\}$ and, in particular, it contains at least one lattice point other than $p_3$ and $p_4$. By central symmetry around its center $\left(\frac{a}2,-\frac{b}2,1\right)$, $\conv(q_2p_3q_1p_4)$ must contain lattice points in both triangles $\conv(q_2p_3p_4)\subset T_1$ and $\conv(q_1p_3p_4)\subset T_2$.

\begin{figure}[htb]
\includegraphics[width=.9\textwidth]{Whites}
\caption{The parallelograms $\conv(p_1p_2q_3q_4)$ and $\conv(q_2q_1p_4p_3)$. The red points are lattice points.}
\label{fig:corner}
\end{figure}
\giulia{ditto}

%\paco{Alternative arguments: one can explicitly compute coordinates of lattice points in $T_1$ and $T_2$, or one can say that the isomorphism $T_{a,b} \cong T_{a^{-1},b}$ of White tetrahedra exchanges the edges at $\{z=0\}$ and at $\{z=1\}$.}

\end{proof}

@@

We now turn our attention to $d=2$, in order to prove \Cref{thm:cayley}. A triangulation of $\cayley(P,Q)\subset \R^3$ consists of tetrahedra of types $(1,3)$, $(2,2)$ and $(3,1)$, where the type denotes how many vertices they have in $P$ and in $Q$. Empty tetrahedra of types $(1,3)$ or $(3,1)$, which are Cayley sums of a triangle in $P$ and a point in $Q$, or viceversa, are automatically unimodular. The case that we need to study are therefore tetrahedra of type $(2,2)$, which are Cayley sums of a segment $p\subset P$ and a segment $q\subset Q$.
\giulia{fix this}
\paco{done, but check it}
\giulia{just changed small things}
The following lemma, whose proof we postpone to \Cref{sec:the_lemma}, is crucial to understand how to unimodularly cover these tetrahedra. 
We use the following conventions: if $a, b$ are points, we denote by $[a,b]$ and $(a,b)$ respectively the closed and open segments with endpoints $a,b$. Given a segment $s=[a,b]$, we denote the vector $\vec s:= b-a$ and  the line spanned by $\vec s$ by $\vecline s$.

@@

\begin{figure}[htb]
\scalebox{.75}{\input{strips.pdf_t}}
\caption{The strips  of Lemma \ref{lemma:cayley}}
\label{fig:strips}
\end{figure}
\giulia{added picture, we may not want it}

@@

\begin{figure}[htb]
\includegraphics[width=.9\textwidth]{flip}
\caption{$[u,q_2]$ intersects $\conv(p_1,p_2,q_1)$}
\label{fig:flip}
\end{figure}

@@

Let us now show how to derive \Cref{coro:prismatoid,coro:width1} from this theorem.
\emph{Prismatoids} were defined in~\cite{Santos-hirsch} as polytopes whose vertices all lie in two parallel facets. In particular, a \emph{lattice prismatoid} is any $d$-polytope $SL(\Z,d)$-equivalent to one of the form
\[
\conv(Q_1\times\{0\} \cup Q_2 \times \{k\}),
\]
where $Q_1,Q_2$ are lattice $(d-1)$-polytopes and $k\in \Z_{>0}$. This is almost a generalization of Cayley sums, which would be the case $k=1$, except the definition of prismatoid requires $Q_1$ and $Q_2$ to be full-dimensional, while the Cayley sum only requires this for $Q_1+Q_2$. \giulia{I didn't like "this is a generalization except...}

@@

We let $H_1=\{f_q(x) \leq 0\}$ and $H_2=\{f_q(x) \geq 0\}$; similarly let $V_1=\{f_p(x) \leq 0\}$ and $V_2=\{f_p(x) \geq 0\}$. \giulia{we need to make these closed, so that the case $P$ one dimensional fits into this same proof}
In the figures, we draw $p$ as a vertical segment and $q$ as a horizontal one, so that $H_i \cap V_j$ are the four quadrants.

@@

\begin{figure}
\includegraphics[scale=.3]{setup.png}
\caption{Setup for \Cref{lemma:cayley}}
\end{figure}

\begin{proof}[Proof of \Cref{lemma:cayley}]
Suppose by contradiction that there is no lattice point as described in the lemma. In particular, no lattice point on the boundary of $Q$ can be in the interior of the strip $q + \vecline p$.  Thus the boundary of $Q$ contains two primitive segments which each have one vertex on each side of the strip $q + \vecline p$; we will call these  $b=[b_1, b_2], t=[t_1, t_2]$, with $b$ and $t$ crossing the strip in $H_1$ and $H_2$ respectively and the convention that $f_p(b_2) >f_p(b_1)$ and $f_p(t_2) >f_p(t_1)$. This readily implies

\begin{gather}
\label{eq:widthq}
\begin{array}{cc}
f_p(t_1) \leq f_p(q_1), &
f_p(t_2) \geq f_p(q_2), \\
f_p(b_1) \leq f_p(q_1), &
f_p(b_2) \geq f_p(q_2).
\end{array}
\end{gather}

>>>>

\begin{abstract}
We show that the following classes of lattice polytopes have unimodular covers, in dimension three: the class of parallelepipeds, the class of centrally symmetric polytopes, and the class of Cayley sums $\cayley(P,Q)$ where the normal fan of $Q$ refines that of $P$. This improves results of Beck et al.~(2018) and Haase et al.~(2008) where the last two classes were shown to be IDP.
\end{abstract}

@@

Oda \cite{Oda1997} posed several questions regarding smoothness and the IDP property for lattice polytopes.
%
Following \cite{HaaseHof, Tsuchiya}, we say that a pair $(P, Q)$ of lattice polytopes has the integer decomposition property, or that \emph{the pair $(P,Q)$ is IDP}, if 
\begin{align*}
\label{eq:mixedIDP}
(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
\end{align*}
%
A lattice polytope $Q$ is called \emph{smooth} if it is simple and the primitive edge directions at every vertex form a linear basis for the lattice; equivalently, if the projective toric variety defined by the normal fan of $Q$ is smooth. 
The following versions of Oda's questions are now considered conjectures~\cite{HNPS2008,mfo2007}, and they are open even in dimension three:
\begin{conjecture}
\label{conj:Oda}
\begin{enumerate}
\item 
\label{itm:smoothIDP}
(Related to problems 2 and 5 in \cite{Oda1997})
Every smooth lattice polytope is IDP.
\item 
\label{itm:mixedIDP}
(Related to problems 1, 3, 4, 6 in \cite{Oda1997}) Every pair $(P,Q)$ of lattice polytopes with $Q$ smooth and the normal fan of $Q$ refining that of $P$ is IDP.
%\begin{align}
%\label{eq:mixedIDP}
%(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
%\end{align}
\end{enumerate}
\end{conjecture}

When the normal fan of a polytope $Q$ refines that of another polytope $P$, as in the second conjecture, we say that $P$ \emph{is a weak Minkowski summand of $Q$}, since this is equivalent to the existence of a polytope $P'$ such that $P+P' = k Q$ for some dilation constant $k>0$\paco{did not find this in the google books version of \cite{Grunbaum}; found it in Postnikov-Reiner-Williams, Tsuchiya, a survey by Rekha Thomas, and other places. Perhaps we can leave it wothout reference}. 
This property has the following algebraic implication for the projective toric variety $X_Q$: $P$ is a weak Minkowski summand of $Q$ if and only if the Cartier divisor defined by $P$ on  $X_Q$ is \emph{numerically effective}, or ``nef'' (see~\cite[Cor.~6.2.15, Prop.~6.3.12]{CLS}, but observe that what we here call  ``weak Minkowski summand'' is simply called ``Minkowski summand'' there). 
%\paco{added this explanation}
%
%\begin{definition}
%Following \cite{HaaseHof, Tsuchiya} we say that the pair $(P, Q)$ has the integer decomposition property, or \emph{the pair $(P,Q)$ is IDP}, if 
%\begin{align*}
%(P+Q) \cap \Z^d = P \cap \Z^d + Q \cap \Z^d.
%\end{align*}
%\end{definition}
% 
%In this language,  \Cref{conj:Oda}\eqref{itm:mixedIDP} can be rephrased as: if $P$ is a weak Minkowski summand of a smooth polytope $Q$ then the pair $(P,Q)$ is IDP.

@@

The two-dimensional case of \Cref{conj:Oda}\eqref{itm:mixedIDP} is known to hold, with three different proofs by Fakhruddin~\cite{Fakhruddin}, Ogata~\cite{Ogata} and Haase et al.~\cite{HNPS2008}. This last one actually shows that smoothness of $Q$ is not needed. In dimension three, however, the conjecture fails without the smoothness assumption. Indeed, if we let $P=Q$ be any non-unimodular \emph{empty tetrahedron}, then $P$ is obviously a weak Minkowski summand of $Q$ but the pair $(P.Q)$ is not IDP. By an empty tetrahedron we mean a lattice tetrahedron containing no lattice points other than its vertices (see the proof of \Cref{lemma:corner} for a classification of them).
\giulia{I didn't like "details on them" and changed it to  "classification of empty tetrahedra". Is this ok?}
\paco{ok. But left it as "classification of them"}

An alternative approach to \Cref{conj:Oda}\eqref{itm:mixedIDP} is via Cayley sums, which we discuss in  \Cref{sec:cayley}. 
Recall that the \emph{Cayley sum} of two lattice polytopes $Q_1,Q_2\subset \R^d$ is the lattice polytope
\[
\cayley(Q_1,Q_2) := \conv(Q_1\times\{0\} \cup Q_2 \times \{1\}) \subset \R^3.
\]
We normally require $\cayley(P,Q)$ to be full-dimensional (otherwise we can delete coordinates) but this does not need $Q_1$ and $Q_2$ to be full-dimensional. It only requires the linear subspaces parallel to them to span $\R^d$.

@@

This has the following two corollaries, also proved in \Cref{sec:cayley}.
A  \emph{prismatoid} is a polytope whose vertices all lie in two parallel facets. 
A polytope has width $1$ if its vertices lie in two \emph{consecutive} parallel lattice hyperplanes. Observe that this is the same as being ($SL(\Z,d)$-equivalent to) a Cayley sum.

@@

For each $i \in [4]$, let $T_i=\conv(q_i, p_j, p_k, p_l)$, with $\{i,j,k,l\}=[4]$; we call these $T_i$ the \emph{corner tetrahedra} of $C(T)$. Together with $T$ they triangulate $C(T)$.
\end{definition}

Modulo an affine transformation, the situation of $T$ and $C(T)$ is exactly that of the regular tetrahedron inscribed in a cube; see \Cref{fig:circumscribed_parall}. 
%
\begin{figure}[htb]
\includegraphics[scale=.25]{circumscribed_parall}
\caption{In red we have a tetrahedron $T$, in black its circumscribed parallelepiped $C(T)$, and in blue the corner simplex $T_4$.}
\label{fig:circumscribed_parall}
\end{figure}

@@

See \Cref{fig:corner}. Observe that $u+v=p_1+p_2=q_3+q_4$.
Now, this implies that the quadrilateral $\conv(p_1q_4p_2q_3)$ contains a fundamental domain for the lattice $\Z^2\times\{0\}$. Hence, its translate $\conv(q_2p_3q_1p_4)$ contains a fundamental domain for $\Z^2\times\{1\}$ and, in particular, it contains at least one lattice point other than $p_3$ and $p_4$. By central symmetry around its center $\left(\frac{a}2,-\frac{b}2,1\right)$, $\conv(q_2p_3q_1p_4)$ must contain lattice points in both triangles $\conv(q_2p_3p_4)\subset T_1$ and $\conv(q_1p_3p_4)\subset T_2$.
%
\giulia{ditto}
\paco{not sure about this one. It looks like the red dots lie on edges of $C(T)$}
\begin{figure}[htb]
\includegraphics[scale=.25]{Whites}
\caption{The parallelograms $\conv(p_1p_2q_3q_4)$ and $\conv(q_2q_1p_4p_3)$. The large red dots are lattice points.}
\label{fig:corner}
\end{figure}
%\paco{Alternative arguments: one can explicitly compute coordinates of lattice points in $T_1$ and $T_2$, or one can say that the isomorphism $T_{a,b} \cong T_{a^{-1},b}$ of White tetrahedra exchanges the edges at $\{z=0\}$ and at $\{z=1\}$.}
\end{proof}

@@

We now turn our attention to $d=2$, in order to prove \Cref{thm:cayley}. A triangulation of $\cayley(P,Q)\subset \R^3$ consists of tetrahedra of types $(1,3)$, $(2,2)$ and $(3,1)$, where the type denotes how many vertices they have in $P$ and in $Q$. Empty tetrahedra of types $(1,3)$ or $(3,1)$, which are Cayley sums of a triangle in $P$ and a point in $Q$, or viceversa, are automatically unimodular. The case that we need to study are therefore tetrahedra of type $(2,2)$, which are Cayley sums of a segment $p\subset P$ and a segment $q\subset Q$.
The following lemma, whose proof we postpone to \Cref{sec:the_lemma}, is crucial to understand how to unimodularly cover these tetrahedra. 
We use the following conventions: if $a, b$ are points, we denote by $[a,b]$ and $(a,b)$ respectively the closed and open segments with endpoints $a,b$. Given a segment $s=[a,b]$, we denote the vector $\vec s:= b-a$ and  the line spanned by $\vec s$ by $\vecline s$.

@@

%\giulia{added picture, we may not want it}
%\paco{looks ok}
\begin{figure}[htb]
\scalebox{.75}{\input{strips.pdf_t}}
\caption{The strips  of Lemma \ref{lemma:cayley}}
\label{fig:strips}
\end{figure}

@@

\begin{figure}[htb]
\includegraphics[scale=.3]{flip}
\caption{$[u,q_2]$ intersects $\conv(p_1,p_2,q_1)$}
\label{fig:flip}
\end{figure}

@@

Let us now show how to derive \Cref{coro:prismatoid,coro:width1} from this theorem.
\emph{Prismatoids} were defined in~\cite{Santos-hirsch} as polytopes whose vertices all lie in two parallel facets. In particular, a \emph{lattice prismatoid} is any $d$-polytope $SL(\Z,d)$-equivalent to one of the form
\[
\conv(Q_1\times\{0\} \cup Q_2 \times \{k\}),
\]
where $Q_1,Q_2$ are lattice $(d-1)$-polytopes and $k\in \Z_{>0}$. This is almost a generalization of Cayley sums, which would be the case $k=1$, except the definition of prismatoid requires $Q_1$ and $Q_2$ to be full-dimensional, while the Cayley sum only requires this for $Q_1+Q_2$.

@@

We let $H_1=\{f_q(x) \leq 0\}$ and $H_2=\{f_q(x) \geq 0\}$; similarly let $V_1=\{f_p(x) \leq 0\}$ and $V_2=\{f_p(x) \geq 0\}$. 
\giulia{we need to make these closed, so that the case $P$ one dimensional fits into this same proof}\paco{ok}
In the figures, we draw $p$ as a vertical segment and $q$ as a horizontal one, so that $H_i \cap V_j$ are the four quadrants. 
See \Cref{fig:setup}.
%
\begin{figure}[htb]
\includegraphics[scale=.3]{setup.png}
\caption{Setup for the proof of \Cref{lemma:cayley}}
\label{fig:setup}
\end{figure}

@@

\begin{proof}[Proof of \Cref{lemma:cayley}]
Suppose by contradiction that there is no lattice point as described in the lemma. In particular, no lattice point on the boundary of $Q$ can be in the interior of the strip $q + \vecline p$.  Thus the boundary of $Q$ contains two primitive segments which each have one vertex on each side of the strip $q + \vecline p$; we will call these  $b=[b_1, b_2], t=[t_1, t_2]$, with $b$ and $t$ crossing the strip in $H_1$ and $H_2$ respectively and the convention that $f_p(b_2) >f_p(b_1)$ and $f_p(t_2) >f_p(t_1)$. This readily implies 
\begin{gather}
\label{eq:widthq}
\begin{array}{cc}
f_p(t_1) \leq f_p(q_1), &
f_p(t_2) \geq f_p(q_2), \\
f_p(b_1) \leq f_p(q_1), &
f_p(b_2) \geq f_p(q_2).
\end{array}
\end{gather}
